---
title: "Simulating data with PH assumptions starting with a life table"
output: html_notebook
---


[Demography 213 - Simulating Data with Proportional Hazards Assumption Starting with a Life Table](http://courses.demog.berkeley.edu/mason213F15/Week10/welcome.pdf)

We first begin by reading in the data. 

```{r}
df = read.table("Data/bltper_5x1.txt", skip = 2, header = TRUE, stringsAsFactors = FALSE)
```

We can then print out the top few rows to make sure that it was inputted correctly:

```{r}
head(df)
```

**lx** is defined as survivorship. 



```{r}
source("snorkHMD.r")
df = cleanHMD(df)

dfu = df[df$Year == "2005",]
dfu$width <- dfu$age.upper - dfu$age.lower + 1
```

```{r}
cdf = 1 - dfu$lx/dfu$lx[1]
plot(dfu$age.lower, cdf , type = "s")
```



```{r}
u = runif(n=1)
u = 0.654

lifetime = df$age.lower[sum(cdf < u) + 1]

plot(dfu$age.lower, cdf , type = "s")
arrows(x0 = 0, x1 = lifetime, y1 = u, y0 = u, col='blue', lwd=3 )
arrows(x0 = lifetime, x1 = lifetime, y1 = 0, y0 = u, col='blue', lwd=3 )
```


```{r}
N = 1e5
lf = runif(n=N) # Generates probabilities btwn 0 and 1
head(lf) 
```

<!-- A brief intermission into how `cut()` works.  -->
<!-- ```{r} -->
<!-- Z <- stats::rnorm(10000) -->
<!-- table(cut(Z, breaks = -6:6)) -->
<!-- sum(table(cut(Z, breaks = -6:6, labels = FALSE))) -->
<!-- ``` -->



```{r}
lives = cut(lf, breaks=c(cdf, 1), labels = dfu$age.lower)
head(lives)
```


```{r}
cbind(table(lives), dfu$dx)
```



```{r}
lx_hat = c(N,N - cumsum(table(lives)))
lx_hat = lx_hat[-length(lx_hat)]
names(lx_hat) = dfu$age.lower

lx_hat
```



```{r}
plot( lx_hat ~ names(lx_hat), type = "s", lwd = 4, 
      col = rgb(red = 0, green = 0, blue = 1, alpha = 0.5)  )
lines(dfu$age.lower, dfu$lx, type = "s", lty = 2, lwd = 2, col = "red")
title("Lifetime table and simulated lifetime")
legend(x = "bottomleft", lty = 2:1, lwd = 2:1, col = c("red", "blue"),
       legend = c("lx US-05", "simulated"), bty = "n")
```



```{r}
plot(dfu$age.lower, cdf, type = "s", lty = 2)
lines(x = dfu$age.lower, y = cdf, type = 'b', lty = 1, lwd = 2, col = 'violet', cex = 0.5)
```


```{r}
cdf = 1 - dfu$lx/dfu$lx[1]

lf = runif(n = 10000)
lives_linear = approx(y = dfu$age.lower, x = cdf, xout = lf )

lives = cut(lf, breaks = c(cdf, 1), labels = dfu$age.lower )
plot(
  y = sort(as.numeric(as.character(lives))),
  x = (1:length(lives)) / length(lives),
  xlab = "percentile", cex = 0.3)
points(
  y = sort(lives_linear$y), pch = 2, cex=0.3, col= "red",
  x = (1:length(sort(lives_linear$y))) / length(lives_linear$y)
)
title("Discrete vs piecewise linear lifetimes")
```



```{r}
lambda = 0.05
ranexp = -log(runif(n=10000)) / lambda
mean(ranexp)
1 / lambda
```

```{r}
plot(sort(ranexp, decreasing = TRUE))
lines(sort(rexp(n=10000, rate = lambda), decreasing = TRUE), col = 'red', lwd = 3)
```

```{r}
qqplot(ranexp, rexp(n = 10000, rate = lambda))
```


```{r}
rise = diff(log(dfu$lx))
run = diff(dfu$age.lower)
hazard = -rise / run
dotchart(hazard, labels = dfu$Age, main = "Hazard Rate by age category")
```


Note the constant hazard rate

```{r}
S_t = function(t, lambda){
  exp(-lambda * t)
}

plot(1:200, S_t(1:200, 0.0070548269), type = "l")
```



```{r}
lxCinverse = function(u, life_tab=dfu){
  survival = c(with(life_tab, lx / lx[1]), 0)
  survival.age = c(life_tab$age.lower, 110)
  
  res = approx(
    y = survival.age,
    x = log(survival + 0.000001),
    xout = log(u)
  )
  return(res$y)
}

with(dfu, plot(age.lower ~ I(lx / lx[1]), type = "s"))
curve(lxCinverse, from = 0, to = 1, n = 10000, lwd = 2, add = TRUE, col = "purple")
```


```{r}
summary(lives_linear$y)
summary(lxCinverse(runif(10000), dfu))
```


```{r}
lives_0 = lxCinverse(u = runif(n = 10000), life_tab = dfu)
dfv = dfu
dfv$lx = dfv$lx^(exp(-0.25))

lives_1 = lxCinverse(u = runif(n = 10000), life_tab = dfv)

dset = data.frame(
                lives = c(lives_0, lives_1), 
                veg = c( rep(0, length(lives_0)), rep(1, length(lives_1) ) )
                )
```


```{r}
library(survival)

dset$depvar = Surv(
  time  = rep(0, nrow(dset)),
  time2 = dset$lives,
  event = rep(1, nrow(dset))
)

cox_res = coxph(depvar ~ veg, data = dset)
summary(cox_res)
```



```{r}
survival = c((dfu$lx / dfu$lx[1]),0)
survival_age = c(dfu$age.lower, 124)
survival_veg = survival^exp(-0.25)

lives_0 = approx(y = survival_age,
                 x = survival,
                 xout = runif(n=10000))$y

lives_1 = approx(y = survival_age,
                 x = survival_veg,
                 xout = runif(n=10000))$y

dset = data.frame(
                lives = c(lives_0, lives_1), 
                veg = c( rep(0, length(lives_0)), rep(1, length(lives_1) ) )
                )

dset$depvar = Surv(
  time  = rep(0, nrow(dset)),
  time2 = dset$lives,
  event = rep(1, nrow(dset))
)

cox_res = coxph(depvar ~ veg, data = dset)
summary(cox_res)
```





# Part II
## Messing with the assumptions underlying the Cox PH model



```{r}
df = read.table("Data/bltper_5x1.txt", skip = 2, header = TRUE, stringsAsFactors = FALSE)
source("snorkHMD.r")
df = cleanHMD(df)
dfu = df[df$Year == "2005",]
```



```{r}
# lx inverse continuous piecewise exponential with Xbeta
lxicpexb = function(Xbeta, life_tab = dfu){
  if(missing(Xbeta)) {
    stop("Xbeta is required and the whole point of the function.")
  }
  
  Xbeta = as.vector(Xbeta)
  
  survival = c(life_tab$lx, 0) / life_tab$lx[1]
  survival = log(survival + 1e-6)
  
  ages = c(life_tab$age.lower, 115)
  
  survivals = outer(survival, exp(Xbeta), "*")
  
  u = runif(n = length(Xbeta))
  res = NULL
  
  for(i in 1:length(u)){
    res = c(res,
            approx(x = survivals[,i] ,
                   y = ages ,
                   xout = log(u[i]))$y )
  }
  return(res)
}
```


```{r}
lf_base = lxicpexb(Xbeta = rep(0, 10000))

mean(lf_base)
```

```{r}
mean(dfu$ex[1]) # Expectancy at Birth
```

```{r}
mean(lf_base[lf_base > 50]) - 50
```

```{r}
dfu$ex[dfu$age.lower == 50]
```

```{r}
mean(lf_base[lf_base > 90]) - 90
```

```{r}
dfu$ex[dfu$age.lower == 90]
```


```{r}
library(survival)
N = 2500

X = data.frame(charm = rnorm(N),
               virtue = runif(N) < 0.5 ,
               angst = rnorm(N))

beta = c(0.1, -0.2, 0.3)
Xbeta = as.matrix(X) %*% beta
names(beta) = c("charm", "virtue", "angst")

dset = data.frame(X, 
                  depvar = Surv(time  = rep(0, N),
                                time2 = lxicpexb(Xbeta = Xbeta, life_tab = dfu),
                                event = rep(1, N))
                  )
```


```{r}
m1 = coxph(depvar ~ charm + virtue + angst, data = dset)
summary(m1)
```


```{r}
Xbeta = as.matrix(X) %*% beta
t2_org = lxicpexb(life_tab = dfu, Xbeta = Xbeta)
t2_mod = log((rank(t2_org) + 1)^0.75) + trunc(t2_org / 25)

plot(t2_mod ~ t2_org, cex = 0.6)
```


```{r}
dset = data.frame(X,
                  depvar_org = Surv(time  = rep(0, N),
                                    time2 = t2_org,
                                    event = rep(1, N)),
                  depvar_mod = Surv(time  = rep(0, N),
                                    time2 = t2_mod,
                                    event = rep(1, N))
                  
                  )
res_org = coxph(depvar_org ~ charm + virtue + angst, data = dset)
summary(res_org)
```

```{r}
res_mod = coxph(depvar_mod ~ charm + virtue + angst, data = dset)
summary(res_mod)
```

```{r}
coefficients(res_org) - coefficients(res_mod)
```


PH model is robust against shifts in time, as it is estimating the shift on the hazard rate at every value of t. So alterations in the baseline hazard rate have no effect on the estimate of the shift between groups.


```{r}
plot(basehaz(res_org), pch = 1, col = 'blue', cex = 0.5, main = "Original")
```


```{r}
plot(basehaz(res_mod), pch = 1, col = 'blue', cex = 0.5, main = "Modified")
```




